---
title: "Stat 331 Group Project"
author: "Tevan Kim, Will Ton, Eli Newman, Trycia Vong"
format: 
  html:
    embed-resources: true
    code-tools: true
    toc: true
editor: source
execute: 
  error: true
  echo: true
  message: false
  warning: false
---

```{r setup}
library(ggplot2)
library(tidyverse)
library(janitor)

gdp <- read_csv("GDP_data.csv", col_types = cols(.default = "c"))
co2 <- read_csv("CO2_Data.csv", col_types = cols(.default = "c"))
```

## Combine data into long format and combine

```{r}
gpd_long <- gdp |> 
  pivot_longer(cols = -country,
               names_to = "year",
               values_to = "GDP")


co2_long <- co2 |> 
  pivot_longer(cols = -country,
               names_to = "year",
               values_to = "co2_emissions")


combined_data <- full_join(gpd_long, co2_long, by = c("country", "year")) |> 
  clean_names()
```

## Introduction

\[Description of data\].

The data we are exploring is the GDP per capita of every country in the world from 1800 projected until 2100 in 2017 international dollars, and the CO2 emissions per capita of every country in the world from 1800 until 2022 in tonnes per capita.

https://www.iea.org/commentaries/the-relationship-between-growth-in-gdp-and-co2-has-loosened-it-needs-to-be-cut-completely


**Hypothesis:** Based on our initial research, we hypothesize that most countries on this list will have a diverging pattern between CO2 emissions and GDP per capita. This, we posit, is due to the fact that more countries have begun adopting clean power generation more widely that previous decades and phasing out fossil fuels. Therefore, GDP per capita continues to grow without having to burn so much CO2.



```{r}

```

## Data Cleaning

```{r}
# First, determine if there is a pattern in missing data
combined_data |> 
  filter(if_any(.cols = everything(),
                .fns = is.na)) |> 
  summarize(na_obs = n(),
            across(.cols = country:co2_emissions,
                   ~ any(is.na(.)))) |> 
  kable()

# Then, remove all observation which contain NA
combined_clean <- combined_data |> 
  filter(!if_any(.cols = everything(),
                .fns = is.na))
```

Discuss any data cleaning process and decisions that may impact the analysis.

## Modeling the Relationship between \_\_\_ and \_\_\_\_

## Cross-Validation

## Conclusion

**Citations**
