---
title: "Stat 331 Group Project"
author: "Tevan Kim, Will Ton, Eli Newman, Trycia Vong"
format: 
  html:
    embed-resources: true
    code-tools: true
    toc: true
editor: source
execute: 
  error: true
  echo: true
  message: false
  warning: false
---

```{r setup}
library(ggplot2)
library(tidyverse)
library(janitor)
library(knitr)

gdp <- read_csv("GDP_data.csv", col_types = cols(.default = "c"))
co2 <- read_csv("CO2_Data.csv", col_types = cols(.default = "c"))
```

## Combine data into long format and combine

```{r}
gpd_long <- gdp |> 
  pivot_longer(cols = -country,
               names_to = "year",
               values_to = "GDP")


co2_long <- co2 |> 
  pivot_longer(cols = -country,
               names_to = "year",
               values_to = "co2_emissions")


combined_data <- full_join(gpd_long, co2_long, by = c("country", "year")) |> 
  clean_names()
```

## Introduction

\[Description of data\].

The data we are exploring is the GDP per capita of every country in the world from 1800 projected until 2100 in 2017 international dollars, and the CO2 emissions per capita of every country in the world from 1800 until 2022 in tonnes per capita.

https://www.iea.org/commentaries/the-relationship-between-growth-in-gdp-and-co2-has-loosened-it-needs-to-be-cut-completely

**Hypothesis:** Based on our initial research, we hypothesize that most countries on this list will have a diverging pattern between CO2 emissions and GDP per capita. This, we posit, is due to the fact that more countries have begun adopting clean power generation more widely that previous decades and phasing out fossil fuels. Therefore, GDP per capita continues to grow without having to burn so much CO2.

```{r}

```

## Data Cleaning

By analyzing the data below, it was determined that all countries were missing CO2 emissions data beyond 2022. Therefore this study will only include data from years 1800 to 2022.

Additionally, it was determined that the only countries missing CO2 emissions data before 2022 were Monaco and San Marino, so those countries will be dropped from this analysis. Similarly, Liechtenstein is missing much of its GDP data before 2022, so it will also be dropped from the analysis. In total, 15,879 obeservations were dropped.

The plot below shows all countries in the dataset missing either CO2 emissions data or GDP data before 2023.

```{r}
#| code-fold: true

# Group Countries into larger regions
europe <- c(
  "Albania", "Andorra", "Austria", "Belarus", "Belgium",
  "Bosnia and Herzegovina", "Bulgaria", "Croatia", "Cyprus",
  "Czech Republic", "Denmark", "Estonia", "Finland", "France",
  "Germany", "Greece", "Hungary", "Iceland", "Ireland", "Italy",
  "Kosovo", "Latvia", "Liechtenstein", "Lithuania", "Luxembourg",
  "Malta", "Moldova", "Monaco", "Montenegro", "Netherlands",
  "North Macedonia", "Norway", "Poland", "Portugal", "Romania",
  "Russia", "San Marino", "Serbia", "Slovakia", "Slovenia",
  "Spain", "Sweden", "Switzerland", "Ukraine", "United Kingdom",
  "Vatican City","UK","Slovak Republic"
)

africa <- c(
  "Algeria","Angola","Benin","Botswana","Burkina Faso","Burundi",
  "Cabo Verde","Cameroon","Central African Republic","Chad",
  "Comoros","Congo, Republic of","Congo, Democratic Republic of",
  "Côte d'Ivoire","Djibouti","Egypt","Equatorial Guinea","Eritrea",
  "Eswatini","Ethiopia","Gabon","Gambia","Ghana","Guinea",
  "Guinea-Bissau","Kenya","Lesotho","Liberia","Libya","Madagascar",
  "Malawi","Mali","Mauritania","Mauritius","Morocco","Mozambique",
  "Namibia","Niger","Nigeria","Rwanda","São Tomé and Príncipe",
  "Senegal","Seychelles","Sierra Leone","Somalia","South Africa",
  "South Sudan","Sudan","Tanzania","Togo","Tunisia","Uganda","Zambia","Zimbabwe",
  "Cote d'Ivoire","Congo, Dem. Rep.","Cape Verde","Sao Tome and Principe","Congo, Rep."
)

asia <- c(
  "Afghanistan","Armenia","Azerbaijan","Bahrain","Bangladesh","Bhutan",
  "Brunei","Cambodia","China","Cyprus","Georgia","India","Indonesia",
  "Iran","Iraq","Israel","Japan","Jordan","Kazakhstan","Kuwait","Kyrgyzstan",
  "Laos","Lebanon","Malaysia","Maldives","Mongolia","Myanmar","Nepal",
  "North Korea","Oman","Pakistan","Palestine","Philippines","Qatar",
  "Saudi Arabia","Singapore","South Korea","Sri Lanka","Syria","Tajikistan",
  "Thailand","Timor-Leste","Turkey","Turkmenistan","UAE","Uzbekistan","Vietnam","Yemen",
  "Hong Kong, China","Kyrgyz Republic","Lao","Taiwan"
)

north_america <- c(
  "Antigua and Barbuda","Bahamas","Barbados","Belize","Canada",
  "Costa Rica","Cuba","Dominica","Dominican Republic","El Salvador",
  "Grenada","Guatemala","Haiti","Honduras","Jamaica","Mexico",
  "Nicaragua","Panama","Saint Kitts and Nevis","Saint Lucia",
  "Saint Vincent and the Grenadines","Trinidad and Tobago","United States","USA"
)

south_america <- c(
  "Argentina","Bolivia","Brazil","Chile","Colombia","Ecuador",
  "Guyana","Paraguay","Peru","Suriname","Uruguay","Venezuela",
  "St. Kitts and Nevis","St. Lucia","St. Vincent and the Grenadines"
)

oceania <- c(
  "Australia","Fiji","Kiribati","Marshall Islands","Micronesia",
  "Nauru","New Zealand","Palau","Papua New Guinea","Samoa",
  "Solomon Islands","Tonga","Tuvalu","Vanuatu","Micronesia, Fed. Sts."
)
```

```{r}
#| warning: false
#| message: false
# First, determine if there is a pattern in missing data
combined_data |> 
  filter(if_any(.cols = everything(),
                .fns = is.na)) |> 
  summarize(na_obs = n(),
            across(.cols = country:co2_emissions,
                   ~ any(is.na(.)))) |> 
  kable()

combined_data |> 
  mutate(year = as.numeric(year),
         country = fct(country),
#         region = fct_collapse(.f = country,
#                               "Europe"        = europe,
#                               "Africa"        = africa,
#                               "Asia"          = asia,
#                               "North America" = north_america,
#                               "South America" = south_america,
#                               "Oceania"       = oceania),
         missing_gdp = is.na(gdp),
         missing_emissions = is.na(co2_emissions)) |> 
  group_by(country, year) |> 
  summarize(prop_missing_gdp = sum(missing_gdp)/n(),
            prop_missing_emissions = sum(missing_emissions)/n()) |> 
  filter(prop_missing_gdp != 0 | prop_missing_emissions != 0,
         !year >= 2023) |> 
  pivot_longer(
    cols      = c(prop_missing_gdp, prop_missing_emissions),
    names_to  = "datatype",
    values_to = "prop_missing") |> 
  ggplot(aes(y = prop_missing,
             x = year,
             fill = datatype)) +
  geom_col(position = position_dodge(width = 0.8),
           width    = 0.7) +
  facet_wrap(~ country) +
  labs(title = "Countries with missing data before 2023",
       subtitle = "Proportion of Missing Data",
       y = "",
       x = "Year",
       fill = "Missing")

# Then, remove all observation which contain NA, and all observations beyond 2022.
combined_clean <- combined_data |> 
  filter(!if_any(.cols = everything(),
                .fns = is.na),
         !year >= 2023)
```

## Modeling the Relationship between GDP per capita  and CO2 emission per capita

Written Components:
Present and discuss visualizations of the relationship between the variables.
Describe the statistical method used – linear regression.
Present the estimated regression model (in notation).
Interpret the linear regression coefficients (in context).
Describe the fit of the regression model (both in table and written format).

2.1: Data Visualization 
  Plot 1: the relationship between the two quantitative variables we are looking at. (aggregate values over all years     for each country)
```{r}
combined_clean |> 
  mutate(gdp = as.numeric(gdp),
         co2_emissions = as.numeric(co2_emissions)) |> 
  group_by(year) |> 
  summarise(total_gdp = sum(gdp, na.rm = TRUE),
            total_co2 = sum(co2_emissions, na.rm = TRUE)) |> 
  ggplot(aes(x = total_gdp, y = total_co2)) +
  geom_point() +
  labs(title = "Aggregate GDP vs CO₂ Emissions by Year",
       x = "Total GDP per capita",
       y = "Total CO₂ Emissions per capita") +
  theme_minimal()
```
  
  
  Plot 2: how the relationship from plot 1 changes over time(animated plot)
  
2.2: Linear Regression
  fit a simple linear regression model to the data. To perform your linear regression, you will likely want to summarize    your data so you have one x value and one y value per country. You may do this by averaging (or taking the median)      across all years, averaging across a subset of years, or picking a year of particular interest ( do the same thing as   in plot #1). Make sure you discuss your decision in your written report.
  
2.3: Model Fit
  Make a nicely formatted table that presents the following:
  - The variance in the response values. (A)
  - The variance in the fitted values from your regression model. (B)
  - The variance in the residuals from your regression model.
  - The model R2 (B / A)




## Cross-Validation

## Conclusion

**Citations**
