---
title: "Stat 331 Group Project"
author: "Tevan Kim, Will Ton, Eli Newman, Trycia Vong"
format: 
  html:
    embed-resources: true
    code-tools: true
    toc: true
editor: source
execute: 
  error: true
  echo: true
  message: false
  warning: false
---

```{r setup}
library(ggplot2)
library(tidyverse)
library(janitor)

gdp <- read_csv("GDP_data.csv", col_types = cols(.default = "c"))
co2 <- read_csv("CO2_Data.csv", col_types = cols(.default = "c"))
```

## Combine data into long format and combine

```{r}
gpd_long <- gdp |> 
  pivot_longer(cols = -country,
               names_to = "year",
               values_to = "GDP")


co2_long <- co2 |> 
  pivot_longer(cols = -country,
               names_to = "year",
               values_to = "co2_emissions")


combined_data <- full_join(gpd_long, co2_long, by = c("country", "year")) |> 
  clean_names()
```

## Introduction

\[Description of data\].

**Hypothesis:** We expect that GDP and CO2 emission (both per capita) are proportional because countries with more resources. . .

```{r}

```

## Data Cleaning

```{r}
# First, determine if there is a pattern in missing data
combined_data |> 
  filter(if_any(.cols = everything(),
                .fns = is.na)) |> 
  summarize(na_obs = n(),
            across(.cols = country:co2_emissions,
                   ~ any(is.na(.)))) |> 
  kable()

# Then, remove all observation which contain NA
combined_clean <- combined_data |> 
  filter(!if_any(.cols = everything(),
                .fns = is.na))
```

Discuss any data cleaning process and decisions that may impact the analysis.

## Modeling the Relationship between \_\_\_ and \_\_\_\_

## Cross-Validation

## Conclusion

**Citations**
