---
title: "The Relationship of GDP per Capita and CO2 Emissions"
author: "Tevan Kim, Will Ton, Eli Newman, Trycia Vong"
format: 
  html:
    embed-resources: true
    code-tools: true
    code-fold: true
    toc: true
editor: source
execute: 
  error: true
  echo: true
  message: false
  warning: false
---

```{r setup}
library(ggplot2)
library(tidyverse)
library(janitor)
library(knitr)
library(gganimate)
library(gifski)
library(broom)
library(kableExtra)

gdp <- read_csv("GDP_data.csv", col_types = cols(.default = "c"))
co2 <- read_csv("CO2_Data.csv", col_types = cols(.default = "c"))
```



```{r}
# Combine data into long format and combine
gpd_long <- gdp |> 
  pivot_longer(cols = -country,
               names_to = "year",
               values_to = "GDP")


co2_long <- co2 |> 
  pivot_longer(cols = -country,
               names_to = "year",
               values_to = "co2_emissions")


combined_data <- full_join(gpd_long, co2_long, by = c("country", "year")) |> 
  clean_names()
```

## Introduction

  The data we are exploring is the GDP per capita of every country in the world from 1800 projected until 2100 in 2017 international dollars, and the CO2 emissions per capita of every country in the world from 1800 until 2022 in tonnes per capita.

Understanding the relationship between economic growth and environmental impact is essential in today's society. In this report we explore the association between gross domestic product (GDP) per capita and carbon dioxide (CO₂) emissions per capita across the world over time. By analyzing these variables, we aim to look at whether higher levels of income are linked to greater environmental costs, and to what extent economic development correlates with carbon output. Understanding this trend could help influence policy making into the future as we aim to curb our carbon emissions and preserve our planet. 

**Hypothesis:** Based on our initial research, we expect most countries to show a positive (direct) linear relationship between GDP and CO2 emissions over time, with a potential decline in emissions relative to GDP in recent years. This, we hypothesize, is due to the fact that more countries have begun adopting clean power generation more widely that previous decades and phasing out fossil fuels. Therefore, GDP per capita continues to grow without having to burn so much CO2.

## Data Cleaning

  By analyzing the data below, it was determined that all countries were missing CO2 emissions data beyond 2022. Therefore this study will only include data from years 1800 to 2022 (this drops 15,210 observations).

Additionally, it was determined that the only countries missing CO2 emissions data before 2022 were Monaco and San Marino, so those countries will be dropped from this analysis. Similarly, Liechtenstein is missing much of its GDP data before 2022, so it will also be dropped from the analysis. These countries totaled 669 observations, so in total 15,879 observations were dropped.

For simplicity, we have decided to summarize our data by averaging the GDP per capita and CO2 emissions for all years by country.

```{r}
#| warning: false
#| message: false
#| code-fold: true
#| echo: true
#| results: false

# First, determine if there is a pattern in missing data
combined_data |> 
  filter(if_any(.cols = everything(),
                .fns = is.na)) |> 
  summarize(na_obs = n(),
            across(.cols = country:co2_emissions,
                   ~ any(is.na(.)))) |> 
  kable()

dc_plot <- combined_data |> 
  mutate(year = as.numeric(year),
         country = fct(country),
#         region = fct_collapse(.f = country,
#                               "Europe"        = europe,
#                               "Africa"        = africa,
#                               "Asia"          = asia,
#                               "North America" = north_america,
#                               "South America" = south_america,
#                               "Oceania"       = oceania),
         missing_gdp = is.na(gdp),
         missing_emissions = is.na(co2_emissions)) |> 
  group_by(country, year) |> 
  summarize(prop_missing_gdp = sum(missing_gdp)/n(),
            prop_missing_emissions = sum(missing_emissions)/n()) |> 
  filter(prop_missing_gdp != 0 | prop_missing_emissions != 0,
         !year >= 2023) |> 
  pivot_longer(
    cols      = c(prop_missing_gdp, prop_missing_emissions),
    names_to  = "datatype",
    values_to = "prop_missing") |> 
  ggplot(aes(y = prop_missing,
             x = year,
             fill = datatype)) +
  geom_col(position = position_dodge(width = 0.8),
           width    = 0.7) +
  facet_wrap(~ country) +
  labs(title = "Countries with missing data before 2023",
       subtitle = "Proportion of Missing Data",
       y = "",
       x = "Year",
       fill = "Missing")

combined_data |> 
  filter(year >= 2023) |> 
  summarize("obs_over_2023" = n()) |> 
  kable()

# Then, remove all observation which contain NA, and all observations beyond 2022.
combined_clean <- combined_data |> 
  filter(!if_any(.cols = everything(),
                .fns = is.na),
         !year >= 2023)
```

## Modeling the Relationship between GDP per capita and CO2 emission per capita

  From our findings, GDP per capita and CO2 emissions have a positive direct relationship; an increase in GDP tends to result in an increase in CO2 emissions. At first this relationship looked non-linear, so a log transformation was applied to the GDP and CO2 variables. This transformation made the trend much more linear. The model we used to describe the trend of the data was linear regression. This was done by taking an average for both GDP and CO2 emissions over all years and performing linear regression calculations over those values. The R-squared value of 0.76 means that there is a strong trend in similarity between the two variables, and it fits a linear model well. The sigma is the standard deviation of the model. In this context, it represents how much the data deviates from the linear model, or how many outliers exist away from the trend. Our std.deviation of about 1 is quite good given the range of values over which the study is conducted. Overall, the model seems to fit quite well with the data and suggests there does exist a relationship between GDP per capita and CO2 emissions per year.

### 2.1: Data Visualization
We start by cleaning our data to get only the key values of interest: GDP and CO2 emissions for each year and for each country. This allows us to easily transition into plotting our data as a nice visual representation as shown below.

```{r}
combined_clean |> 
  mutate(
  gdp = case_when(
    grepl("k$", gdp, ignore.case = TRUE) ~ as.numeric(sub("k", "", gdp, ignore.case = TRUE)) * 1000,
    TRUE ~ as.numeric(gsub(",", "", gdp))
  )) |>
  mutate(co2_emissions = as.numeric(gsub(",", "", co2_emissions))) |> 
  group_by(country) |> 
  summarise(avg_gdp = mean(gdp, na.rm = TRUE),
            avg_co2 = mean(co2_emissions, na.rm = TRUE)) |> 
  filter(!is.na(avg_gdp) & !is.na(avg_co2)) |> 
  filter(avg_gdp > 0, avg_co2 > 0) |>  # required for log
  mutate(log_gdp = log(avg_gdp), log_co2 = log(avg_co2)) |> #log transformation
  ggplot(aes(x = log_gdp, y = log_co2)) +
  geom_point(color = "steelblue", size = 3) +
  labs(
    title = "Average CO2 Emissions vs. Average GDP per Country",
    x = "Log(Average GDP Per Capita)",
    y = "Log(Average CO₂ Emissions Per Capita)") +
  theme_minimal()
```
Figure 1: The relationship between log of avg GDP per capita and log of avg CO2 per capita (Aggregate values over all years for each country). Very linear relationship between the two.

```{r}
animated_data <- combined_clean |> 
    mutate(
  gdp = case_when(
    grepl("k$", gdp, ignore.case = TRUE) ~ as.numeric(sub("k", "", gdp, ignore.case = TRUE)) * 1000,
    TRUE ~ as.numeric(gsub(",", "", gdp))
  )) |>
  mutate(co2_emissions = as.numeric(gsub(",", "", co2_emissions))) |> 
  mutate(year = as.integer(year)) |> 
  group_by(country, year) |> 
  summarise(avg_gdp = mean(gdp, na.rm = TRUE),
            avg_co2 = mean(co2_emissions, na.rm = TRUE)) |> 
  filter(!is.na(avg_gdp) & !is.na(avg_co2)) |>
  filter(avg_gdp > 0, avg_co2 > 0) |>  # required for log
  mutate(log_gdp = log(avg_gdp), log_co2 = log(avg_co2)) #log transformation

animated_plot <- ggplot(animated_data, aes(x = log_gdp, y = log_co2)) +
  geom_point(color = "steelblue", size = 3, alpha = 0.7) +
  labs(title = "Average CO2 Emissions vs. Average GDP per Country",
       subtitle = "Year: {frame_time}",
       x = "Log(Average GDP Per Capita)",
       y = "Log(Average CO₂ Emissions Per Capita)") +
  theme_minimal() +
  transition_time(year) +
  ease_aes('linear')


# animate(animated_plot, renderer = gifski_renderer(), width = 800, height = 600, fps = 10)
# anim_save("avg_gdp_vs_co2_over_time.gif", animation = last_animation())

animate(animated_plot, nframes = 100, fps = 10)

```
Figure 2: Animated version of Fig. 1, over the full timescale of the Gapminder Dataset (1800-2022).

### 2.2: Linear Regression

Once fully cleaned, we then fit our data to a linear model to test our hypothesis of a positive trend over time. This is accomplished by guessing a linear fit to our average data points and showing that the trendline agrees with the change in the data over time.

```{r}
#Average gdp and co2 per country over all years
#used grepl and gsub from chatgpt, https://chatgpt.com/share/6837b5d4-9aa0-8009-9faf-83f213beced7
my_model <- combined_clean |>
  group_by(country) |> 
  #turns 12.3k to 12300 or 12k to 12000
  mutate(
  gdp = case_when(
    grepl("k$", gdp, ignore.case = TRUE) ~ as.numeric(sub("k", "", gdp, ignore.case = TRUE)) * 1000,
    TRUE ~ as.numeric(gsub(",", "", gdp))
  )) |>
  mutate(co2_emissions = as.numeric(co2_emissions)) |>
  summarise(avg_gdp_all_years = mean(gdp, na.rm = TRUE),
            avg_co2_all_years = mean(co2_emissions, na.rm = TRUE)) |>
  filter(avg_gdp_all_years > 0, avg_co2_all_years > 0) |>  # required for log
  mutate(log_gdp = log(avg_gdp_all_years), log_co2 = log(avg_co2_all_years)) #linear transformation

regression_data <- lm(log_co2 ~ log_gdp, data = my_model)

tidy(regression_data) |> kable()
broom::glance(regression_data) |>  kable()

# Create a dataframe with fitted values and residuals
residual_data <- my_model |>
  mutate(
    fitted = fitted(regression_data),
    residuals = resid(regression_data)
  )

# Plot residuals vs. fitted values to check linearity
#Plot made from chatgpt, https://chatgpt.com/share/68436ab0-9880-8009-8962-fbc0aeeff57f
ggplot(residual_data, aes(x = fitted, y = residuals)) +
  geom_point(color = "darkred", size = 3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  labs(
    title = "Residual Plot",
    subtitle = "Residuals",
    x = expression("Fitted Values (Predicted " * CO[2] *" Emissions)"),
    y = ""
  ) +
  theme_minimal()

```
Figure 3: residuals vs fitted plot of CO2 emissions. Some fanning is present, but is not too bad. 


### 2.3: Model Fit

  An R-squared value of 0.84 means that approximately 84% of the variability in average CO2 emissions can be explained by our model, and therefore by average GDP. This suggests the quality of our model is decent, though 24% of the variability in average CO2 emissions is due to other factors not captured in our model.

```{r}
aug <- augment(regression_data) |> 
  summarize(var_response = var(log_co2),
            var_fitted = var(.fitted),
            var_resid = var(.resid),
            R_squared = var_fitted/var_response)

aug |> 
  kable(col.names  = c("Response (A)", "Fitted (B)", "Residuals", "(B/A)"),
        align     = "c") |> 
  kable_styling(full_width = F,
                bootstrap_options = c("striped")) |> 
  add_header_above(c("Variance" = 3, "R\u00B2" = 1)) |> 
  add_header_above(c("Model Fit" = 4),
                   font_size = "x-large")
```

## Cross-Validation
To prove our model can work for a variety of test data sets, we performed k-fold cross validation to prove this. This was done by randomly assigning data points into k number of groups, where k, in this study, is 10. From there, we reserve one of the groups as a test dataset while the nine others are used to train a linear model with which we test against the test data. After performing tests k times, we averaged the R2 values to see if the R2 of our original linear model is close. For this study, there was only a difference of about 0.01, which means the linear model is successfully validated.

```{r}

oneIteration <- function (dataset, k){
  test_data <- filter(dataset, fold == k)
  train_data <- filter(dataset, fold != k)
  
  trained_fit <- lm(log_co2 ~ log_gdp, data = train_data)
  
  test_preds <- predict(trained_fit, newdata = test_data)
  
  residual <- test_data$log_co2 - test_preds
  
  return(1 - (var(residual))/(var(test_data$log_co2)))
}

n <- nrow(my_model)
set.seed(57482)

k <- 10
vec <- 1:k
my_model <- my_model |> 
  mutate(fold = sample(rep_len(1:k, length.out = n),
                       size = n)) 

cv <- map_dbl(vec, ~ oneIteration(my_model, .x))
mean(cv)


```
```{r}
#Expression() from geeksforgeeks.org
df <- data.frame(k_fold = vec, r_2 = cv)

df |>
  ggplot(aes(x = k_fold, y = r_2)) +
  geom_point() +
  labs(title = expression("Distribution of R" ^ 2 * " Values vs. K Folds"),
       subtitle = expression("R" ^ 2 * " Values"),
       x = "K Fold Number",
       y = "") +
  geom_hline(yintercept = mean(cv),
             linetype = "dashed",
             size = 0.5,
             color = "black") +
  geom_hline(yintercept = aug$R_squared,
             linetype = "dashed",
             size = 0.5,
             color = "blue") +
  annotate(
    "text",
    x = 9.5,
    y = mean(cv) + 0.01,
    label = "k-fold average",
    color = "black"
  ) +
  annotate(
    "text",
    x = 9.5,
    y = aug$R_squared + 0.01,
    label = "original model fit",
    color = "blue"
  ) +
  scale_x_continuous(limits = c(1, 10), breaks = seq(1, 10, 1)) +
  theme_minimal()
```
Figure 4: K fold plot showing that the k-fold average is very close to the original model fit, meaning no over or under fitting has occured. 

## Conclusion

  In this report, we have explored the relationship between GDP per capita of every country in the world from 1800 until 2022 in 2017 international dollars, and the CO2 emissions per capita of every country in the world from 1800 until 2022. We hypothesized that there'd be a direct linear relationship with an inverse linear trend in recent years due to cleaner energy use in order to combat pollution. Our linear regression model showed a strong positive relationship with an R squared value of 0.84, which suggests that approximately 84% of the variability in average CO2 emissions can be explained by our model, and therefore by average GDP. However, we did not observe an inverse trend in more recent years. This could be due to the averaging of data across all years per country, which may have obscured more recent trends. It may also indicate that while cleaner technologies are emerging, their impact has not yet outweighed increasing economic activity and energy consumption in many countries. 
  Through our k-fold cross validation we found an average R squared of 0.82 with relatively low variance between folds. This suggests that the model's predictive performance indicates good generalization. In conclusion, from our data we can see that the relationship between GDP per capita and CO2 emissions has always been a positive relationship showing that an increase in GDP tends to result in an increase in CO2 emissions from the years 1800 to 2022. We can see that supports our initial hypothesis of a "direct linear relationship", but does not support the initial hypothesis of having an inverse linear trend because of countries switching to cleaner power generation. 



**Citations**

https://www.iea.org/commentaries/the-relationship-between-growth-in-gdp-and-co2-has-loosened-it-needs-to-be-cut-completely

